---
title: "A Bayesian Alternative to the Wilcoxon Rank Test"
author: "Joachim Reinhardt"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: Rbibliography.bib
vignette: >
  %\VignetteIndexEntry{A Bayesian Alternative to the Wilcoxon Rank Test}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The **wilcox_test** R-package implements a Bayesian Alternative to the Wilcoxon 
Signed Rank test and is intended as an addition to the **BayesianFirstAid** 
package [@baath2014bayesian]. The approach follows [@gelman2014bayesian, p. 97],
who  suggest a three-step methodology for a Bayesian approach to the Wilcoxon 
test. Following this methodology, the **wilcox.test** function first transforms 
the combined input data to their ranks as in the conventional Wilcoxon 
procedure. Then, these ranks are normalized into z-scores, and the difference 
between the two groups is modeled as a normal distribution with uninformative 
priors. This approach has the following advantages: ... 

Note: Consider Mann-Whitney Test! -> test for two independent groups, also
included in wilcox.test function

Gelman (blog post): "the Wilcoxon test has a huge assumption which is that the 
distributions are identical in the two groups. This is a much more important 
assumption than normality."



## Introduction

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## A Bayesian Alternative to the Wilcoxon

### The frequentist Wilcoxon Test



## Literature Overview

[@gelman2011bayesian, p.11]: "Even the Wilcoxon Test is based on assumptions!"

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))

## References
